---
title: "EPIC Orientation 2018: Introduction to R"
date: "Updated: `r format(Sys.Date(), format = '%d %B %Y')`" 
output:
  html_document:
    theme: default
    highlight: zenburn
  md_document: default
  variant: markdown_github
github_document: default
number_sections: true
---


## Overview
1. Importing data
2. R Data Types
3. Working with dataframes
4. Data Manipulation with dplyr 
5. Reshaping data with tidyr
6. Data Visualization using ggplot2

## 1. Importing data
For starters, we will be working with `tv_hurricanes_by_network.csv`, which contains the percentage of TV news that mention Hurricanes Harvey, Irma, Jose, and Maria each day.

Source: https://github.com/fivethirtyeight/data/tree/master/puerto-rico-media

Let's get started.
```{r}
#clean environment
rm(list = ls())

#to get current working directory
getwd() 

#to get help for a function
help("read.csv")

#import csv
df <- read.csv("datasets/tv_hurricanes_by_network.csv")
```

R also has the capability to import other file formats including Stata, SPSS, SAS and Excel files using functions such as `read.spss()`, `read.dta()`, and `read.xport()`. You may wish to explore the `foreign` package to find out more. 

## 2. R Data Types

You will encounter a variety of data types in R:

* #### scalar 
    + a single number (e.g. 3)
* #### vector
    + `numeric <- c(0.1, 2.2. 4.678)`
    + `integer <- c(5, 10, 100)`
    + `character <- c("EPIC", "is", "epic!")`
    + `logical <- c(T, F, T, T)`
    + `factor <- c(1,2,3,4)`
* #### matrices
    + all columns in a matrix must be of the same type (numeric, character, etc.) and the same length.
* #### dataframe
    + more general than a matrix, different columns can have different modes (numeric, character, factor, etc.)
* #### list
    + an ordered collection of objects (components), a list allows you to gather (possibly unrelated) objects under one name (e.g. a        string, a numeric vector, a matrix, and a scalar)  e.g. `mylist <- list("Hello", c(1,2,5.3,6,-2,4), matrix(1:20, nrow=5,ncol=4), 9)`
    + you can also make lists of lists e.g. `mylist <- list(list1, list2, list3)`

Let's examine our data.

```{r}
#see structure of object
str(df)
```

```{r}
#see class of a variable
class(df$Query)
```

R is currently treating the Query variable as nominal, let's change it to a character type.
```{r}
#change Query variable to a character type
df$Query <- as.character(df$Query) 
#see structure
str(df)
```

## 3. Working with Dataframes
Sometimes we deal with large dataframes and we don't always get the luxury of viewing it in its entirety. Instead, we can use the following functions:
```{r}
#see dimensions (in this order: row, column)
dim(df)
#view the first 2 obs
head(df, 2)
#view the last 3 obs
tail(df, 3)
#summary statistics
summary(df)
```

To access specific columns or rows by name or index:
```{r}
#Get the 2nd observation of the Query Variable
df[2, 'Query'] #Method 1
df$Query[2] #Method 2
#Get the 3rd observation of the 5th variable
df[3,5]
#Grab multiple elements (from the 1st & 2nd rows, 4th & 5th columns)
df[c(1,2), c(4,5)]
```


You can also perform functions or use operators on vectors
```{r}
#Find the minimum value of CNN
min(df$CNN) 
#Sum of all obs in BBC.News
sum(df$BBC.News)
#Standard deviation of Fox News
sd(df$FOX.News)
#Multiple CNN by 2
df$CNN.times.2 <- df$CNN*2
#Remove this newly created column
df$CNN.times.2 <- NULL
#Add variables together
df$CNN.plus.BBC <- df$CNN + df$BBC.News
#Convert Date variable from Factor to Date format
df$Date <- as.Date(df$Date, format = "%m/%d/%y")
#Tabulate the Query variable, how many observations are Hurricane Irma?
table(df$Query)
```

There are several ways to rename a column. For instance, if we'd like to rename the "Query"" column to "Hurricane"
```{r}
#view all variable names in df
names(df)
#Method 1: rename the second variable, aka Query
names(df)[2] <- "HURRICANE" #Method 1
#Method 2: rename based on the name
names(df)[names(df)=="HURRICANE"] <- "Hurricane"
#view all variable names in df again
names(df)
#Method 3: use rename() from the dplyr package which we will cover later
```

### Quiz 1: 
Create a numeric variable that assigns the following damage costs to each of the hurricanes. Name this variable `Costs` and use millions as the unit. 

* Harvey: $125Bn
* Maria: $90Bn
* Irma: $64.8Bn
* Jose: $2.8M

##### *“If you look for it, I've got a sneaky feeling you'll find that love actually is all around.” - Prime Minister David (Hugh Grant)*

```{r}
df$Costs[df$Hurricane=="Hurricane Harvey"] <- 125 * 1000
df$Costs[df$Hurricane=="Hurricane Irma"] <- 64.8 * 1000
df$Costs[df$Hurricane=="Hurricane Maria"] <- 90 * 1000
df$Costs[df$Hurricane=="Hurricane Jose"] <- 2.8
```

## 4. Data Manipulation with dplyr 

#### What's dplyr?
dplyr is a powerful package written by the amazing Hadley Wickham. It provides some easy-to-use functions for data analysis and manipulation.

```{r}
#install package
#install.packages("dplyr")
#load this package in your R environment
library(dplyr)
```

Now let's try out some of the most commonly-used functions.

* `select()` - select specific columns 
* `mutate()` - create a new column
* `rename()` - rename columns 
* `filter()` - select rows using a criteria
* `arrange()` - arrange rows
* `group_by()` - group data
* `summarise()` - summarize values

`select()` allows you to select specific columns from your dataframe.

```{r}
#select just the Date and CNN columns and display the first 3 observations
#my R is buggy but you should be able to run just this: head(select(df, Date, CNN))
head(dplyr::select(df, Date, CNN), 3)
```
`mutate()` allows you to create new variables

```{r}
#create a new column called BBC.and.CNN and display just the first 3 rows
head(mutate(df, BBC.and.CNN=BBC.News+CNN), 3)
```
`rename()` allows you to rename variables.
```{r}
#change the variable name BBC.News to just BBC and FOX.News to just FOX
#again, my R is buggy but you should be able to run just this: head(rename(df, c("BBC.News" = "BBC", "FOX.News" = "FOX")),3)
head(plyr::rename(df, c("BBC.News" = "BBC", "FOX.News" = "FOX")),3)
```

`filter()` allows you to select specific observations using conditions
```{r}
#find the first 3 observations where CNN has a value of < 0.1 and MSNBC < 0.02
head(filter(df, CNN < 0.1 & MSNBC < 0.02),3)
```

`arrange()` allows you to rearrange rows 
```{r}
#arrange rows according to Dates (in ascending order), then Hurricane
head(arrange(df, Date, Hurricane), 3)
```

`group_by()` uses a **split-apply-combine** concept. To break it down, we want to **split** the dataframe into groups by some variable you've specified, **apply** a function to these groups, and **combine** them again. Let's try this with the `summarise()` function.

```{r}
#Find out the mean news coverage in CNN by hurricane groups
df %>%
  group_by(Hurricane) %>%
  summarise(mean.coverage = mean(CNN))
```

#### What is this `%>%` symbol?
This is known as a **pipe operator**, which allows you to pipe the output from one function to the input of another function. This saves you from long, nested operations or doing a bunch of assignments. 

Without the pipe operator, you could also execute the previous script with a long, nested operation (too confusing):
```{r}
summarise(group_by(df, Hurricane), mean.coverage = mean(CNN))
```

Or using multiple assignments (too tedious):
```{r}
test <- group_by(df, Hurricane)
test <- summarise(test, mean.coverage = mean(CNN))
```

### Quiz 2: 
Using a combination of the functions we've covered, find out the average coverage of MSNBC for each hurricane, only include hurricanes where damage costs were above 10 million. Remember to use the pipe operator :D

```{r}
df %>%
  filter(Costs > 10) %>%
  group_by(Hurricane) %>%
  summarise(max.coverage = max(MSNBC)) %>%
  arrange(max.coverage)
  
```

Just to make a point, if you **didn't** use the pipe operator, a nested operation would have looked like this (don't do it).
```{r}
arrange(summarise(group_by(filter(df, Costs > 10), Hurricane), max.coverage = max(MSNBC)), max.coverage)
```

## 5. Reshaping data with tidyr 

#### What's tidyr?
The most commonly used functions in this package are `gather()` and `spread()`, which help us reshape our data. Our data currently has 4 separate columns for news coverage, one for each TV station. What if we wanted to restructure our data from a **wide to long** format, so the data of all 4 TV stations can be in a single variable?

```{r}
#install.packages("tidyr")
library(tidyr)
#reshape from wide to long
plot_df <- gather(df, 
        key = "tv.station", #name of the new column of 'old columns to be gathered'
        value = "coverage", #name of the 'gathered values' 
        BBC.News:MSNBC) #columns to be 'gathered'
```

`spread()` is the complement of the above. Try it out!
```{r}
#reshape from long to wide
head(spread(plot_df, tv.station, coverage), 3)
```

## 6. Data visualization with ggplot2 
The `ggplot2` package is based on the **grammar of graphics**, the idea that you can build every graph from the same few components: a data set, a set of geoms (data points), and a coordinate system.

```{r}
#install.packages ("ggplot2")
#install.packages ("scales")
library(ggplot2) 
library(scales)

#view dataframe for plotting 
head(plot_df)

#create density plots  
ggplot(data = plot_df) +
  geom_density(aes(x = coverage, fill = tv.station), alpha = 0.3) +
  xlab("news coverage")  #label the x axis

#create a timeseries of news coverage 
ggplot(data = plot_df) +
  geom_line(aes(x = as.Date(Date, "%m/%d/%y"), y = coverage, 
                group = tv.station, 
                color = tv.station), alpha = 0.7) +
  xlab("date") + 
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("1 week")) 

#we could also have separate plots by tv station by using facet_wrap()
ggplot(data = plot_df) +
  geom_line(aes(x = as.Date(Date, "%m/%d/%y"), y = coverage, 
                group = tv.station, 
                color = tv.station), alpha = 0.5) +
  facet_wrap(~tv.station, ncol = 4) + 
  xlab("date") + 
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("2 weeks")) 

#create histograms of each hurricane in terms of number of observations
ggplot(data = plot_df) +
  geom_bar(aes(x = Hurricane, fill = Hurricane), 
           color = "black",
           stat = "count", #calculates the number of cases in each group
           alpha = 0.3) 

#create scatterplot of news coverage against time by TV Station
ggplot(data = plot_df) +
  geom_point(aes(y = coverage, x = as.Date(Date, "%m/%d/%y"), color = tv.station),
             shape = 16,
             size = 4,
             alpha = 0.3) + 
  xlab("date") + 
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("1 week")) 

```

You may need to overlay plots sometimes, you can do this via layers, remember that they need to share common x and y axes.
```{r}
#overlay scatterplot on timeseries lineplot
ggplot(data = plot_df) +
  geom_line(aes(x = as.Date(Date, "%m/%d/%y"), y = coverage, 
                group = tv.station, 
                color = tv.station), alpha = 0.7) +
  geom_point(aes(y = coverage, x = as.Date(Date, "%m/%d/%y"), color = tv.station),
             shape = 16,
             size = 4,
             alpha = 0.3) + 
  xlab("date") + 
  scale_x_date(labels = date_format("%m/%d"),
               breaks = date_breaks("1 week")) 

```



#### Further Resources
* https://github.com/yixinsun1216/covertoperations_manual/wiki/R-Guide
* http://genomicsclass.github.io/book/pages/dplyr_tutorial.html
* https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html
* https://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html
